## This is the analysis pipeline associated with "Modeling the influence of oceanography and zooplankton dynamics on the abundance and distribution of larval tunas in the tropical central Pacific" by Jaskiel et al.

### Set Working Directory, Install/Load Packages
```{r}
setwd("/Users/Jacob/Documents/Ch_3_Materials/Larval_Tuna_GAMs")
getwd() ### "/Users/Jacob/Documents/Ch_3_Materials/Larval_Tuna_GAMs"

### Installing packages if needed:
#install.packages("corrplot")
#install.packages("statmod")
#install.packages("gratia")
#install.packages("car")
#install.packages("pROC")
#install.packages("lunar")
#install.packages("caret")
#install.packages("glmmTMB")
#install.packages("emmeans")
#install.packages("DHARMa")
#install.packages("multcomp")
#install.packages("multcompView")

### Library packages
library(ggplot2)
library(corrplot)
library(mgcv)
library(statmod)
library(tidyr)
library(dplyr)
library(gratia)
library(car)
library(pROC)
library(lunar)
library(caret)
library(glmmTMB)
library(emmeans)
library(DHARMa)
library(multcomp)
library(multcompView)
```

### Bring in data for individual larvae (freq. of occurrence, length, etc.)
```{r}
individual_data <- read.csv("all_PIPA_larvae.csv")

### Make sure cruise number is a factor and lengths are numeric
individual_data <- individual_data %>%
  mutate(Cruise = as.factor(Cruise)) %>%
  mutate(Total.Length..mm. = as.numeric(Total.Length..mm.))

individual_data_summary <- individual_data %>%
  group_by(Cruise, genus) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(Cruise, genus)

individual_data_station_summary <- individual_data %>%
  group_by(Station, genus) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(Station, genus)
```

### Plot frequency of occurrence for 1mm length bins for both genera
```{r}
### Katsuwonus percent frequency of occurrence per cruise
#png("katsuwonus_length_freq_by_cruise.png", units = "in", width = 6, height = 4, res = 1200)
ggplot(data = filter(individual_data, genus == "Katsuwonus"), 
       aes(x = Total.Length..mm.)) +
  geom_histogram(binwidth = 1, 
                 boundary = 0, 
                 fill = "darkslategray2", 
                 color = "black", 
                 linewidth = 0.3,
                 aes(y = after_stat(count / tapply(count, PANEL, sum)[PANEL] * 100))) +
  facet_wrap(~ Cruise, 
             scales = "free_y", 
             ncol = 3) +
  labs(x = "Total Length (mm)", 
       y = "Frequency (%)") +
  theme_minimal()
#dev.off()

### Thunnus percent frequency of occurrence per cruise
#png("thunnus_length_freq_by_cruise.png", units = "in", width = 6, height = 4, res = 1200)
ggplot(data = filter(individual_data, genus == "Thunnus"),
       aes(x = Total.Length..mm.)) + 
  geom_histogram(binwidth = 1, 
                 boundary = 0, 
                 fill = "goldenrod1", 
                 color = "black", 
                 linewidth = 0.3,
                 aes(y = after_stat(count / tapply(count, PANEL, sum)[PANEL] * 100))) +
  facet_wrap(~ Cruise,
             scales = "free_y", 
             ncol = 3) +
  labs(x = "Total Length (mm)", 
       y = "Frequency (%)") +
  theme_minimal()
#dev.off()

### Combine
length_freq_data <- individual_data %>%
  filter(genus %in% c("Katsuwonus", "Thunnus")) %>%
  filter(!is.na(Total.Length..mm.)) %>%
  mutate(length_bin = floor(Total.Length..mm.)) %>%
  group_by(genus, Cruise, length_bin) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(genus, Cruise) %>%
  mutate(pct = count / sum(count) * 100) %>%
  ungroup()

#png("combined_length_freq.png", units = "in", width = 6, height = 4, res = 1200)
ggplot(length_freq_data, 
       aes(x = length_bin + 0.5, 
           y = pct, 
           fill = genus)) +
  geom_col(position = position_dodge2(width = 1, preserve = "single"), 
           color = "black", 
           linewidth = 0.3, 
           width = 1) +
  facet_wrap(~ Cruise, 
             scales = "free_y", 
             ncol = 3) +
  scale_fill_manual(values = c("Katsuwonus" = "darkslategray2", 
                               "Thunnus" = "goldenrod1")) +
  labs(x = "Total Length (mm)", 
       y = "Frequency (%)", 
       fill = "Genus") +
  theme_minimal()
#dev.off()

table(individual_data$genus[!is.na(individual_data$Total.Length..mm.)])
```

### Reading in and making sure data are correct class
```{r}
station_data <- read.csv("Fish_plankton_tuna.csv")
str(station_data)

### Making variables the correct class
station_data$Jar.. <- as.factor(station_data$Jar..)
station_data$Date <- as.Date(station_data$Date, format="%d-%b-%y")
station_data$Cruise <- as.factor(station_data$Cruise)
station_data$Year <- as.factor(station_data$Year)
station_data$Day.or.Night <- as.factor(station_data$Day.or.Night)
station_data$ADCP_velo <- as.numeric(station_data$ADCP_velo)
station_data$Water.Depth <- as.numeric(station_data$Water.Depth)
station_data$Tow.Depth..m. <- as.numeric(station_data$Tow.Depth..m.)
station_data$risen.or.set <- as.factor(station_data$risen.or.set)
station_data$Distance..km..to.nearest.island <- as.numeric(station_data$Distance..km..to.nearest.island)
station_data$What.is.the.nearest.island. <- as.factor(station_data$What.is.the.nearest.island.)
station_data$Tow.Type <- as.factor(station_data$Tow.Type)
station_data$Region <- as.factor(station_data$Region)
station_data$Current <- as.factor(station_data$Current)
station_data$Tow.Length..m. <- as.numeric(station_data$Tow.Length..m.)
station_data$Tow.Volume..m3. <- as.numeric(station_data$Tow.Volume..m3.)
station_data$Zpl.Density..ml.m3. <- as.numeric(station_data$Zpl.Density..ml.m3.)
station_data$Tuna.Density..ml.m3. <- as.numeric(station_data$Tuna.Density..ml.m3.)

### The rest are numeric so let's just do this:
station_data[34:148] <- lapply(station_data[34:148], as.numeric)

### Let's look now
str(station_data)

### Looks good, ecxept let's remove the one sample with a bunch of phytoplankton goop (S261-034-NT)
station_data <- station_data %>% filter(Jar.. != "S261-034-NT")

### Make densities in ind./1000m^3
station_data$Skipjack.Density.1000m3 <- station_data$Skipjack.Density * 1000
station_data$Thunnus.Density.1000m3 <- station_data$Thunnus.Density * 1000
station_data$non.tuna.density.1000m3 <- station_data$non.tuna.density * 1000
station_data$Zpl.Density..ml.1000m3 <- station_data$Zpl.Density..ml.m3. * 1000

### Doing lunar angle instead of % illuminance (note: lunar.phase returns radians hence the transformation below)
station_data$lunar_phase_angle <- lunar.phase(station_data$Date, name = FALSE) * 180 / pi
station_data$day_since_newmoon <- station_data$lunar_phase_angle / 360 * 29.53

### Should range 0-360
range(station_data$lunar_phase_angle, na.rm = TRUE)
range(station_data$day_since_fullmoon, na.rm = TRUE)

### Make deployments w/ > 0 larvae -> 1, none = 0
station_data <- station_data %>%
mutate(Skipjack_pa = as.integer(Skipjack > 0), Thunnus_pa = as.integer(Thunnus > 0))

### Make time 0-24
station_data$hour <- floor(station_data$Time / 100) + (station_data$Time %% 100) / 60
station_data$hour[station_data$hour == 24] <- 0

### See number of NAs (this is from Caroline):
na_table_tidyverse <- station_data %>%
  dplyr::summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "Column Name", 
               values_to = "NA Count")

print(na_table_tidyverse)
```

### Mean Density of larval genera across cruises
```{r}
density_summary <- station_data %>%
  filter(!is.na(Skipjack.Density) | !is.na(Thunnus.Density)) %>%
  pivot_longer(cols = c(Skipjack.Density, Thunnus.Density), 
               names_to = "genus", 
               values_to = "density_m3") %>%
  mutate(genus = case_when(genus == "Skipjack.Density" ~ "Katsuwonus", 
                           genus == "Thunnus.Density"  ~ "Thunnus", 
                           TRUE ~ NA_character_),
         genus = factor(genus, levels = c("Katsuwonus", "Thunnus")),
         density_per_1000m3 = density_m3 * 1000) %>%
  filter(!is.na(genus)) %>%
  group_by(Cruise, genus) %>%
  summarise(mean_density = mean(density_per_1000m3, na.rm = TRUE),
            sd_density = sd(density_per_1000m3, na.rm = TRUE),
            n_stations = n(),
            n_stations_present = sum(density_per_1000m3 > 0, na.rm = TRUE),
            freq_occurrence = (n_stations_present / n_stations) * 100,
            se_density = sd_density / sqrt(n_stations),
            min_density = min(density_per_1000m3, na.rm = TRUE),
            max_density = max(density_per_1000m3, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(Cruise = factor(Cruise))

#png("tuna_density_by_cruise.png", units = "in", width = 6, height = 5, res = 1200)
ggplot(density_summary,
       aes(x = Cruise, 
           y = mean_density, 
           fill = genus)) +
  geom_col(position = position_dodge(width = 0.75), 
           width = 0.70, 
           color = "black", 
           linewidth = 0.4) +
  geom_errorbar(aes(ymin = mean_density - se_density, 
                    ymax = mean_density + se_density),
                    position = position_dodge(width = 0.75), 
                    width = 0.45, 
                    linewidth = 0.4) +
  scale_fill_manual(values = c("Katsuwonus" = "darkslategray2", 
                               "Thunnus" = "goldenrod1")) +
  labs(x = "Cruise", 
       y = "Mean density (larvae/1,000 m³)", 
       fill = "Genus") +
  theme_minimal(base_size = 14) +
  theme_minimal()
#dev.off()

### Non-zero stations only
density_summary_nonzero <- station_data %>%
  filter(!is.na(Skipjack.Density) | !is.na(Thunnus.Density)) %>%
  pivot_longer(cols = c(Skipjack.Density, Thunnus.Density), 
               names_to = "genus", 
               values_to = "density_m3") %>%
  mutate(genus = case_when(genus == "Skipjack.Density" ~ "Katsuwonus", 
                           genus == "Thunnus.Density"  ~ "Thunnus", 
                           TRUE ~ NA_character_),
         genus = factor(genus, levels = c("Katsuwonus", "Thunnus")),
         density_per_1000m3 = density_m3 * 1000) %>%
  filter(!is.na(genus), density_per_1000m3 > 0) %>%
  group_by(Cruise, genus) %>%
  summarise(mean_density = mean(density_per_1000m3, na.rm = TRUE), 
            sd_density = sd(density_per_1000m3, na.rm = TRUE), 
            n_stations = n(),
            se_density = sd_density / sqrt(n_stations),
            min_density = min(density_per_1000m3, na.rm = TRUE),
            median_density = median(density_per_1000m3, na.rm = TRUE),
            max_density = max(density_per_1000m3, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(Cruise = factor(Cruise))

### Nonzero density by cruise
#png("nonzero_tuna_density_by_cruise.png", units = "in", width = 6, height = 5, res = 1200)
ggplot(density_summary_nonzero,
       aes(x = Cruise, 
           y = mean_density, 
           fill = genus)) +
  geom_col(position = position_dodge(width = 0.75), 
           width = 0.70, 
           color = "black", 
           linewidth = 0.4) +
  geom_errorbar(aes(ymin = mean_density - se_density, 
                    ymax = mean_density + se_density),
                    position = position_dodge(width = 0.75),
                    width = 0.45,
                    linewidth = 0.4) +
  scale_fill_manual(values = c("Katsuwonus" = "darkslategray2", 
                               "Thunnus" = "goldenrod1")) +
  labs(x = "Cruise",
       y = "Mean density (larvae/1,000 m³)",
       fill = "Genus") +
  theme_minimal(base_size = 14) +
  theme_minimal()
#dev.off()
```

### Boxplot Version
```{r}
### Making the data long format
density_long <- station_data %>%
  pivot_longer(cols = c(Skipjack.Density, Thunnus.Density), 
               names_to = "genus", 
               values_to = "density_m3") %>%
  mutate(genus = case_when(genus == "Skipjack.Density" ~ "Katsuwonus", 
                           genus == "Thunnus.Density"  ~ "Thunnus", 
                           TRUE ~ NA_character_),
         genus = factor(genus, levels = c("Katsuwonus", "Thunnus")), 
         Cruise = factor(Cruise),
         density_per_1000m3 = density_m3 * 1000) %>%
  filter(!is.na(genus), !is.na(density_per_1000m3))

density_long_nonzero <- density_long %>%
  filter(!is.na(density_per_1000m3), density_per_1000m3 > 0)

nonzero_means <- density_long_nonzero %>%
  group_by(Cruise, genus) %>%
  summarise(mean_density = mean(density_per_1000m3, na.rm = TRUE), .groups = "drop")

#png("nonzero_density_by_cruise_boxplot_final.png", units = "in", width = 6, height = 5, res = 1200)
ggplot(density_long_nonzero, 
       aes(x = Cruise, 
           y = density_per_1000m3, 
           fill = genus)) +
  geom_boxplot(aes(group = interaction(Cruise, genus)),
               position = position_dodge(width = 0.75),
               width = 0.70,
               color = "black",
               linewidth = 0.4,
               outlier.shape = 21) +
  geom_point(data = nonzero_means,
             aes(x = Cruise, 
                 y = mean_density, 
                 group = genus),
             position = position_dodge(width = 0.75),
             color = "black",
             size = 2,
             inherit.aes = FALSE) +
  scale_fill_manual(values = c("Katsuwonus" = "darkslategray2",
                               "Thunnus" = "goldenrod1")) +
  scale_y_continuous(limits = c(0, 12)) +
  labs(x = "Cruise",
       y = "Larval Density (ind./1,000 m³)",
       fill = "Genus") +
  theme_minimal(base_size = 14)
#dev.off()
```

###Testing for significant differences between cruises
```{r}
long_data_nonzero <- station_data %>%
  filter(!is.na(Skipjack.Density) | !is.na(Thunnus.Density)) %>%
  pivot_longer(cols = c(Skipjack.Density, Thunnus.Density),
               names_to = "genus",
               values_to = "density_m3") %>%
  mutate(genus = case_when(genus == "Skipjack.Density" ~ "Katsuwonus",
                           genus == "Thunnus.Density" ~ "Thunnus",
                           TRUE ~ NA_character_),
         genus = factor(genus, levels = c("Katsuwonus", "Thunnus")),
         density_per_1000m3 = density_m3 * 1000,
         Cruise = factor(Cruise)) %>%
  filter(!is.na(genus), density_per_1000m3 > 0)

### Kruskal-Wallis 
### Katsuwonus
katsu_density_data_long <- long_data_nonzero %>% filter(genus == "Katsuwonus")
kw_katsu <- kruskal.test(density_per_1000m3 ~ Cruise, data = katsu_density_data_long)
print(kw_katsu)
print(pairwise.wilcox.test(katsu_density_data_long$density_per_1000m3,
                           katsu_density_data_long$Cruise,
                           p.adjust.method = "bonferroni",
                           exact = FALSE))

### Thunnus
thun_density_data_long <- long_data_nonzero %>% filter(genus == "Thunnus")
kw_thun <- kruskal.test(density_per_1000m3 ~ Cruise, data = thun_density_data_long)
print(kw_thun)
print(pairwise.wilcox.test(thun_density_data_long$density_per_1000m3, 
                           thun_density_data_long$Cruise, 
                           p.adjust.method = "bonferroni", 
                           exact = FALSE))
```

### Distribution of variables of interest
```{r}
### Tuna genera
hist(station_data$Skipjack.Density, breaks = 100)
hist(station_data$Thunnus.Density, breaks = 100)

### Environmental parameters
hist(station_data$SST...C., breaks = 100)
hist(station_data$Chla, breaks = 100)
hist(station_data$SSS, breaks = 100)
hist(station_data$Zoop.Biomass..ml., breaks = 100) ### one outlier from prymnesiophytes (phytoplankton) so remove it (S261-034-NT)
hist(station_data$Moon.Phase.., breaks = 100)
hist(station_data$day_since_newmoon, breaks = 100)
hist(station_data$lunar_phase_angle, breaks = 100)
hist(station_data$ADCP_velo, breaks = 100)
hist(station_data$LatDec, breaks = 100)
hist(station_data$LonDec, breaks = 100)
hist(station_data$Time, breaks = 100)

### specific taxa of interest
hist(station_data$Calanoid.Copepod, breaks = 100)
hist(station_data$Chaetognath, breaks = 100)
hist(station_data$Euphausid.Shrimp, breaks = 100)
hist(station_data$Appendicularian, breaks = 100)
hist(station_data$Engraulidae, breaks = 100)
hist(station_data$non.tuna.density, breaks = 100)
hist(station_data$Myctophidae, breaks = 100)

plot(station_data$Chaetognath, station_data$Skipjack.Density)
plot(station_data$Chaetognath, station_data$Thunnus.Density)
```

###Exploratory Correlation Plots
```{r}
numeric_data <- station_data[sapply(station_data, is.numeric)]
numeric_data_reduced <- numeric_data[, c("ADCP_velo", "SST...C.", "Chla", "SSS", "Zoop.Biomass..ml.", "Zpl.Density..ml.m3.", "Tuna.Density..ml.m3.", "Skipjack.Density", "Thunnus.Density", "Calanoid.Copepod", "Chaetognath", "Corycaeus.Copepod", "Cladoceran", "Euphausid.Shrimp", "Salp", "Appendicularian", "Engraulidae", "Exocoetidae", "Hemiramphidae", "Nomeidae", "non.tuna.density")]

cor(numeric_data_reduced, method = "spearman", use = "complete.obs")

### Heatmap
corrplot::corrplot(cor(numeric_data_reduced, method="spearman", use = "complete.obs"))

### only variables used in GAMs
input_vars <- station_data[, c("day_since_newmoon", "Zpl.Density..ml.m3.", "ADCP_velo", "SST...C.", "Chla", "SSS", "LatDec", "hour", "non.tuna.density")]
cor(input_vars, use = "pairwise.complete.obs")
corrplot::corrplot(cor(input_vars, method="spearman", use = "complete.obs"))
```

## GAM for Density 
### Density-based analysis of Skipjack
### Using densities in 1000m^3
```{r}
### Skipjack - density (therefore using binomial family with log link)
gam_skj_dens_tw <- gam(Skipjack.Density.1000m3 ~
                         factor(Year)
                       + s(Thunnus.Density.1000m3, bs ="ts") ### significant (p = 0.000515)
                       + s(Zpl.Density..ml.1000m3, bs = "ts") ### significant (p = 3.31e-05)
                       + s(LatDec, LonDec, bs = "gp", k = 4) ### significant (p < 2e-16)
                       + s(day_since_newmoon, bs = "cc") 
                       + s(ADCP_velo, bs = "ts") ### significant (p < 2e-16)
                       + s(SST...C., bs = "ts") ### significant (p = 0.003282)
                       + s(non.tuna.density.1000m3, bs = "ts") ### significant (p < 2e-16)
                       + s(Chla, bs = "ts") ### significant (p = 0.000405)
                       + s(SSS, bs = "ts") ### significant (p < 2e-16)
                       + s(hour, bs = "cc"), ### significant (p = 8.40e-05)
                       select = TRUE,
                       family = tw(link = "log"),
                       data = station_data,
                       method = "REML")

### Checking performance
summary(gam_skj_dens_tw) 
plot(gam_skj_dens_tw)
gam.check(gam_skj_dens_tw)
AIC(gam_skj_dens_tw) 
gratia::draw(gam_skj_dens_tw)
anova(gam_skj_dens_tw)
concurvity(gam_skj_dens_tw)
vif(gam_skj_dens_tw)
acf(residuals(gam_skj_dens_tw))

### GLM Version for comparison
# glm_skj_dens_tw <- gam(Skipjack.Density.1000m3 ~
#                          factor(Year)
#                        + s(Thunnus.Density.1000m3, bs ="ts") ### significant (p = 0.000515)
#                        + s(Zpl.Density..ml.1000m3, bs = "ts") ### significant (p = 3.31e-05)
#                        + s(LatDec, LonDec, bs = "gp", k = 4) ### significant (p < 2e-16)
#                        + s(day_since_newmoon, bs = "cc")
#                        + s(ADCP_velo, bs = "ts") ### significant (p < 2e-16)
#                        + s(SST...C., bs = "ts") ### significant (p = 0.003282)
#                        + s(non.tuna.density.1000m3, bs = "ts") ### significant (p < 2e-16)
#                        + s(Chla, bs = "ts") ### significant (p = 0.000405)
#                        + s(SSS, bs = "ts") ### significant (p < 2e-16)
#                        + s(hour, bs = "cc"), ### significant (p = 8.40e-05)
#                        select = TRUE,
#                        family = tw(link = "log"),
#                        data = station_data,
#                        method = "REML")

### Checking performance
# summary(glm_skj_dens_tw)
# #plot(glm_skj_dens_tw)
# gam.check(glm_skj_dens_tw)
# AIC(glm_skj_dens_tw)
# gratia::draw(glm_skj_dens_tw)
# anova(glm_skj_dens_tw)
# #concurvity(glm_skj_dens_tw)
# vif(glm_skj_dens_tw)
# acf(residuals(glm_skj_dens_tw))

### Plotting final GAM
### Full Model
#png("Katsu_density_full.png", units = "in", width = 9, height = 9, res = 1200)
gratia::draw(gam_skj_dens_tw, parametric = TRUE, ncol = 3)
#dev.off()

## 5-fold CV
set.seed(54507845)
k <- 5
n <- nrow(station_data)
folds <- sample(rep(1:k, length.out = n))
cv_preds <- numeric(n) ### now holds predicted mean densities

for (i in 1:k) {
  cat("Fold", i, "of", k, "... ")
  train_df <- station_data[folds != i, , drop = FALSE]
  test_df  <- station_data[folds == i, , drop = FALSE]
  gam_fit <- gam(Skipjack.Density.1000m3 ~
                   factor(Year)
                 + s(Thunnus.Density.1000m3, bs ="ts") # significant (p = 0.000515)
                 + s(Zpl.Density..ml.1000m3, bs = "ts") # significant (p = 3.31e-05)
                 + s(LatDec, LonDec, bs = "gp", k = 4) # significant (p < 2e-16)
                 + s(day_since_newmoon, bs = "cc") 
                 + s(ADCP_velo, bs = "ts") # significant (p < 2e-16)
                 + s(SST...C., bs = "ts") # significant (p = 0.003282)
                 + s(non.tuna.density.1000m3, bs = "ts") # significant (p < 2e-16)
                 + s(Chla, bs = "ts") # significant (p = 0.000405)
                 + s(SSS, bs = "ts") # significant (p < 2e-16)
                 + s(hour, bs = "cc"), # significant (p = 8.40e-05)
                 select = TRUE,
                 family = tw(link = "log"),
                 data = train_df,
                 method = "REML")
  cv_preds[folds == i] <- predict(gam_fit,
                                  newdata = test_df,
                                  type = "response")
}

### Fit a GAM to describe predicted vs observed relationship
calib_gam <- gam(station_data$Skipjack.Density.1000m3 ~ s(cv_preds, bs = "ts"),
                 family = tw(link = "log"), 
                 method = "REML")
summary(calib_gam)

### Predict smooth + uncertainty
newdata <- data.frame(cv_preds = seq(0, 120, length.out = 477))
pred <- predict(calib_gam, 
                newdata = newdata, 
                type = "response", 
                se.fit = TRUE)
fit_mean <- pred$fit
ci_lower <- pmax(0, fit_mean - 1.96 * pred$se.fit)
ci_upper <- fit_mean + 1.96 * pred$se.fit

### Plot
#png("Katsu_density_calibration_clipped2.png", units = "in", width = 6, height = 4, res = 1200)
plot(0, 0, 
     type = "n",
     xlim = c(0, 40), 
     ylim = c(0, 80),
     xlab = "Predicted Density", 
     ylab = "Observed Density",
     las = 1, 
     bty = "l")
abline(0, 1, 
       col = "gray50", 
       lty = 2, 
       lwd = 1.6)
polygon(c(newdata$cv_preds, rev(newdata$cv_preds)),
        c(ci_lower, rev(ci_upper)),
        col = rgb(0.78, 0.78, 0.78, 0.35), 
        border = NA)
lines(newdata$cv_preds, 
      fit_mean, 
      col = "black", 
      lwd = 2.6)
points(newdata$cv_preds, 
       station_data$Skipjack.Density.1000m3,
       pch = 16,
       col = rgb(0, 0, 0, 0.25),
       cex = 0.8)
#dev.off()
```

### Density-based analysis of Thunnus
```{r}
gam_thun_dens_tw <- gam(Thunnus.Density.1000m3 ~
                          factor(Year)
                        + s(Skipjack.Density.1000m3, bs = "ts") ### significant (p < 2e-16)
                        + s(Zpl.Density..ml.1000m3, bs = "ts") ### significant (p < 2e-16)
                        + s(non.tuna.density.1000m3, bs = "ts") ### significant (p = 7.53e-07)
                        + s(SST...C., bs = "ts") ### significant (p = 0.00331)
                        + s(Chla, bs = "ts") ### significant (p = 6.50e-07)
                        + s(LatDec, LonDec, bs = "gp", k = 4) ### significant (p = 1.02e-06) 
                        + s(SSS, bs = "ts")
                        + s(hour, bs = "cc")
                        + s(day_since_newmoon, bs = "cc") 
                        + s(ADCP_velo, bs = "ts"),
                        select = TRUE,
                        family = tw(link = "log"),
                        data = station_data,
                        method = "REML")

### Checking performance
summary(gam_thun_dens_tw) 
plot(gam_thun_dens_tw)
gam.check(gam_thun_dens_tw)
AIC(gam_thun_dens_tw) 
gratia::draw(gam_thun_dens_tw)
anova(gam_thun_dens_tw)
concurvity(gam_thun_dens_tw)
vif(gam_thun_dens_tw)
acf(residuals(gam_thun_dens_tw))

### GLM Version for comparison
# glm_thun_dens_tw <- gam(Thunnus.Density.1000m3 ~
#                           factor(Year)
#                         + s(Skipjack.Density.1000m3, bs = "ts") # significant (p < 2e-16)
#                         + s(Zpl.Density..ml.1000m3, bs = "ts") # significant (p < 2e-16)
#                         + s(non.tuna.density.1000m3, bs = "ts") # significant (p = 7.53e-07)
#                         + s(SST...C., bs = "ts") # significant (p = 0.00331)
#                         + s(Chla, bs = "ts") # significant (p = 6.50e-07)
#                         + s(LatDec, LonDec, bs = "gp", k = 4) #significant (p = 1.02e-06)
#                         + s(SSS, bs = "ts")
#                         + s(hour, bs = "cc")
#                         + s(day_since_newmoon, bs = "cc")
#                         + s(ADCP_velo, bs = "ts"),
#                         select = TRUE,
#                         family = tw(link = "log"),
#                         data = station_data,
#                         method = "REML")

### Checking performance
# summary(glm_thun_dens_tw)
# #plot(glm_thun_dens_tw)
# gam.check(glm_thun_dens_tw)
# AIC(glm_thun_dens_tw)
# #gratia::draw(glm_thun_dens_tw)
# anova(glm_thun_dens_tw)
# #concurvity(glm_thun_dens_tw)
# vif(glm_thun_dens_tw)
# acf(residuals(glm_thun_dens_tw))

### Plotting final GAM
### Full Model
#png("Thunnus_density_full.png", units = "in", width = 9, height = 9, res = 1200)
gratia::draw(gam_thun_dens_tw, parametric = TRUE, ncol = 3)
#dev.off()

## 5-fold CV
set.seed(31) 
k <- 5
n <- nrow(station_data)
folds <- sample(rep(1:k, length.out = n))
cv_preds <- numeric(n)          ### now holds predicted mean densities

for (i in 1:k) {
  train_df <- station_data[folds != i, , drop = FALSE]
  test_df  <- station_data[folds == i, , drop = FALSE]
  gam_fit <- gam(Thunnus.Density.1000m3 ~
                   factor(Year)
                 + s(Skipjack.Density.1000m3, bs = "ts") # significant (p < 2e-16)
                 + s(Zpl.Density..ml.1000m3, bs = "ts") # significant (p < 2e-16)
                 + s(non.tuna.density.1000m3, bs = "ts") # significant (p = 7.53e-07)
                 + s(SST...C., bs = "ts") # significant (p = 0.00331)
                 + s(Chla, bs = "ts") # significant (p = 6.50e-07)
                 + s(LatDec, LonDec, bs = "gp", k = 4) #significant (p = 1.02e-06) 
                 + s(SSS, bs = "ts")
                 + s(hour, bs = "cc")
                 + s(day_since_newmoon, bs = "cc") 
                 + s(ADCP_velo, bs = "ts"),
                 select = TRUE,
                 family = tw(link = "log"),
                 data = train_df,
                 method = "REML")
  cv_preds[folds == i] <- predict(gam_fit,
                                  newdata = test_df,
                                  type = "response")
}

### Fit a GAM to describe predicted vs observed relationship
calib_gam <- gam(station_data$Thunnus.Density.1000m3 ~ s(cv_preds, bs = "ts"),
                 family = tw(link = "log"),
                 method = "REML")
summary(calib_gam)

### predict smooth + uncertainty
newdata <- data.frame(cv_preds = seq(0, 40, length.out = 477))
pred <- predict(calib_gam, 
                newdata = newdata, 
                type = "response", 
                se.fit = TRUE)
fit_mean <- pred$fit
ci_lower <- pmax(0, fit_mean - 1.96 * pred$se.fit)
ci_upper <- fit_mean + 1.96 * pred$se.fit

#png("Thunnus_density_calibration_clipped2.png", units = "in", width = 6, height = 4, res = 1200)
plot(0, 0, 
     type = "n",
     xlim = c(0, 10), 
     ylim = c(0, 22),
     xlab = "Predicted Mean Density", 
     ylab = "Observed Mean Density",
     las = 1, 
     bty = "l")
abline(0, 1, 
       col = "gray50", 
       lty = 2, 
       lwd = 1.6)
polygon(c(newdata$cv_preds, rev(newdata$cv_preds)),
        c(ci_lower, rev(ci_upper)),
        col = rgb(0.78, 0.78, 0.78, 0.35), 
        border = NA)
lines(newdata$cv_preds, 
      fit_mean, 
      col = "black", 
      lwd = 2.6)
points(newdata$cv_preds, 
       station_data$Thunnus.Density.1000m3,
       pch = 16,
       col = rgb(0, 0, 0, 0.25),
       cex = 0.8)
#dev.off()
```

## Presence/Absence Models for Supporting Information
### P/A - Skipjack
```{r}
### Skipjack - presence/absence (therefore using binomial family with log link)
gam_skj_pa_binom <- gam(Skipjack_pa ~
                          Thunnus_pa ### incorporating Thunnus presence as a linear parametric predictor (not smooth)
                        + factor(Year)
                        + s(non.tuna.density.1000m3, bs = "ts") ### significant (p = 5.69e-05)
                        + s(SSS, bs = "ts") ### significant (p = 1.54e-05)
                        + s(Zpl.Density..ml.1000m3, bs = "ts") ### significant (p = 0.00115)
                        + s(day_since_newmoon, bs = "cc") 
                        + s(LatDec, LonDec, bs = "gp", k = 4) ### significant (p = 0.00201)
                        + s(Chla, bs = "ts")
                        + s(hour, bs = "cc")
                        + s(SST...C., bs = "ts")
                        + s(ADCP_velo, bs = "ts"),
                        select = TRUE,
                        family = binomial(link = "logit"),
                        data = station_data,
                        method = "REML")

### Checking performance
summary(gam_skj_pa_binom) 
plot(gam_skj_pa_binom)
gam.check(gam_skj_pa_binom)
AIC(gam_skj_pa_binom) 
gratia::draw(gam_skj_pa_binom)
anova(gam_skj_pa_binom)
concurvity(gam_skj_pa_binom)
vif(gam_skj_pa_binom)
acf(residuals(gam_skj_pa_binom))

### GLM Version for comparison
# glm_skj_pa_binom <- gam(Skipjack_pa ~
#                           Thunnus_pa ### incorporating Thunnus presence as a linear parametric predictor (not smooth)
#                         + factor(Year)
#                         + non.tuna.density.1000m3 
#                         + SSS 
#                         + Zpl.Density..ml.1000m3 
#                         + day_since_newmoon
#                         #+ LatDec, LonDec ### this throws an error in the GLM
#                         + Chla
#                         + hour
#                         + SST...C.
#                         + ADCP_velo,
#                         family = binomial(link = "logit"),
#                         data = station_data,
#                         method = "REML")

# summary(glm_skj_pa_binom) 
# #plot(glm_skj_pa_binom)
# #gam.check(glm_skj_pa_binom)
# AIC(glm_skj_pa_binom) 
# gratia::draw(glm_skj_pa_binom)
# anova(glm_skj_pa_binom)
# #concurvity(glm_skj_pa_binom)
# vif(glm_skj_pa_binom)
# acf(residuals(glm_skj_pa_binom))

### Plotting final GAM
### Full
#png("Katsu_PA_full.png", units = "in", width = 9, height = 9, res = 1200)
gratia::draw(gam_skj_pa_binom, parametric = TRUE, ncol = 3)
#dev.off()

## 5-fold CV
set.seed(8)
k <- 5
n <- nrow(station_data)
folds <- sample(rep(1:k, length.out = n))
cv_probs <- numeric(n)
cv_lp <- numeric(n)

for (i in 1:k) {
  train_df <- station_data[folds != i, , drop = FALSE]
  test_df  <- station_data[folds == i, , drop = FALSE]
  gam_fit <- gam(Skipjack_pa ~
                   Thunnus_pa ### incorporating Thunnus presence as a linear parametric predictor (not smooth)
                 + factor(Year)
                 + s(non.tuna.density.1000m3, bs = "ts") ### significant (p = 5.69e-05)
                 + s(SSS, bs = "ts") ### significant (p = 1.54e-05)
                 + s(Zpl.Density..ml.1000m3, bs = "ts") ### significant (p = 0.00115)
                 + s(day_since_newmoon, bs = "cc") 
                 + s(LatDec, LonDec, bs = "gp", k = 4) ### significant (p = 0.00201)
                 + s(Chla, bs = "ts")
                 + s(hour, bs = "cc")
                 + s(SST...C., bs = "ts")
                 + s(ADCP_velo, bs = "ts"),
                 select = TRUE,
                 family = binomial(link = "logit"),
                 data = train_df,
                 method = "REML")
  
  ### Predict on held-out fold
  cv_probs[folds == i] <- predict(gam_fit,
                                  newdata = test_df,
                                  type = "response")
   cv_lp[folds == i] <- predict(gam_fit, 
                                newdata = test_df, 
                                type = "link")
}
observed <- station_data$Skipjack_pa

### Basic summary
mean(cv_probs, na.rm = TRUE) ### 0.2818454
mean(observed) ### 0.2767296

### AUC
roc_obj <- roc(observed, cv_probs, quiet = TRUE)
auc(roc_obj) ### 0.7638

### Calibration plot shows whether the predicted probabilities are well-calibrated (i.e., when the model says 30% chance, do we actually see presence ~30% of the time?)
### Fit the calibration GAM (using out-of-fold predictions)
calib_gam <- gam(station_data$Skipjack_pa ~ s(cv_probs, bs = "ts"),
                 family = binomial(link = "logit"),
                 method = "REML")
summary(calib_gam)

### Predict smooth curve + uncertainty
newdata <- data.frame(cv_probs = seq(0, 1, length.out = 477))
pred <- predict(calib_gam, newdata = newdata, type = "link", se.fit = TRUE)

### convert to probability scale
fit_prob <- plogis(pred$fit)
ci_lower <- plogis(pred$fit - 1.96 * pred$se.fit)
ci_upper <- plogis(pred$fit + 1.96 * pred$se.fit)

### Plot
#png("Katsu_PA_calibration.png", units = "in", width = 6, height = 4, res = 1200)
plot(0, 0, type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "Predicted Probability", 
     ylab = "Observed Probability",
     las = 1, bty = "l", 
     cex.main = 1.1, 
     cex.lab = 1.1)
abline(0, 1, 
       col = "gray50", 
       lty = 2, 
       lwd = 1.6)
polygon(c(newdata$cv_probs, rev(newdata$cv_probs)),
        c(ci_lower, rev(ci_upper)),
        col = rgb(0.78, 0.78, 0.78, 0.35),
        border = NA)
lines(newdata$cv_probs, 
      fit_prob, 
      col = "black", 
      lwd = 2.6)
rug(cv_probs[station_data$Skipjack_pa == 0], 
    side = 1, 
    col = "gray40", 
    lwd = 0.8, 
    ticksize = 0.025)
rug(cv_probs[station_data$Skipjack_pa == 1], 
    side = 3, 
    col = "gray20", 
    lwd = 0.8, 
    ticksize = 0.025)
#dev.off()
```

### Presence/Absence of Thunnus
```{r}
gam_thun_pa_binom <- gam(Thunnus_pa ~
                           Skipjack_pa ### Skipjack presence as a linear parametric predictor (not smooth); significant (p = 0.016439)
                         + factor(Year) ### significant (p = 3.86e-06)
                         + s(Zpl.Density..ml.1000m3, bs = "ts") ### significant (p = 5.09e-05)
                         + s(non.tuna.density.1000m3, bs = "ts") ### significant (p = 0.000563)
                         + s(SST...C., bs = "ts") ### significant (p = 0.003326)
                         + s(LatDec, LonDec, bs = "gp", k = 4) ### significant (p = 0.026190)
                         + s(day_since_newmoon, bs = "cc")
                         + s(Chla, bs = "ts") ### significant (p = 0.000208)
                         + s(ADCP_velo, bs = "ts")
                         + s(hour, bs = "cc")
                         + s(SSS, bs = "ts"),
                         select = TRUE,
                         family = binomial(link = "logit"),
                         data = station_data,
                         method = "REML")

### Checking performance
summary(gam_thun_pa_binom) 
plot(gam_thun_pa_binom)
gam.check(gam_thun_pa_binom)
AIC(gam_thun_pa_binom) 
gratia::draw(gam_thun_pa_binom)
anova(gam_thun_pa_binom)
concurvity(gam_thun_pa_binom)
vif(gam_thun_pa_binom)
acf(residuals(gam_thun_pa_binom))

### GLM Version for comparison
# glm_thun_pa_binom <- gam(Thunnus_pa ~
#                            Skipjack_pa # Skipjack presence as a linear parametric predictor (not smooth); significant (p = 0.016439)
#                          + factor(Year) #significant (p = 3.86e-06)
#                          + s(Zpl.Density..ml.1000m3, bs = "ts") # significant (p = 5.09e-05)
#                          + s(non.tuna.density.1000m3, bs = "ts") # significant (p = 0.000563)
#                          + s(SST...C., bs = "ts") # significant (p = 0.003326)
#                          #+ s(LatDec, LonDec, bs = "gp", k = 4) # significant (p = 0.026190)
#                          + s(day_since_newmoon, bs = "cc")
#                          + s(Chla, bs = "ts") # significant (p = 0.000208)
#                          + s(ADCP_velo, bs = "ts")
#                          + s(hour, bs = "cc")
#                          + s(SSS, bs = "ts"),
#                          select = TRUE,
#                          family = binomial(link = "logit"),
#                          data = station_data,
#                          method = "REML")

### Checking performance
# summary(glm_thun_pa_binom) 
# #plot(glm_thun_pa_binom)
# gam.check(glm_thun_pa_binom)
# AIC(glm_thun_pa_binom) 
# #gratia::draw(glm_thun_pa_binom)
# anova(glm_thun_pa_binom)
# #concurvity(glm_thun_pa_binom)
# vif(glm_thun_pa_binom)
# acf(residuals(glm_thun_pa_binom))

### Plotting final GAM
### Full
#png("Thunnus_PA_full.png", units = "in", width = 9, height = 9, res = 1200)
gratia::draw(gam_thun_pa_binom, parametric = TRUE, ncol = 3)
#dev.off()

## 5-fold CV
set.seed(8)
k <- 5
n <- nrow(station_data)
folds <- sample(rep(1:k, length.out = n))
cv_probs <- numeric(n)
cv_lp <- numeric(n)

for (i in 1:k) {
  train_df <- station_data[folds != i, , drop = FALSE]
  test_df  <- station_data[folds == i, , drop = FALSE]
  gam_fit <- gam(Thunnus_pa ~
                   Skipjack_pa #incorporating Skipjack presence as a linear parametric predictor (not smooth); significant (p = 0.016439)
                 + factor(Year) #significant
                 + s(Zpl.Density..ml.1000m3, bs = "ts") # significant (p = 5.09e-05)
                 + s(non.tuna.density.1000m3, bs = "ts") # significant (p = 0.000563)
                 + s(SST...C., bs = "ts") # significant (p = 0.003326)
                 + s(LatDec, LonDec, bs = "gp", k = 4) # significant (p = 0.026190)
                 + s(day_since_newmoon, bs = "cc")
                 + s(Chla, bs = "ts") # significant (p = 0.000208)
                 + s(ADCP_velo, bs = "ts")
                 + s(hour, bs = "cc")
                 + s(SSS, bs = "ts"),
                 select = TRUE,
                 family = binomial(link = "logit"),
                 data = train_df,
                 method = "REML")
  
  cv_probs[folds == i] <- predict(gam_fit,
                                  newdata = test_df,
                                  type = "response")
   cv_lp[folds == i] <- predict(gam_fit, 
                                newdata = test_df, 
                                type = "link")
}
observed <- station_data$Thunnus_pa

### Basic summary
mean(cv_probs, na.rm = TRUE) ### 0.1988498
mean(observed) ### 0.1949686

### AUC
roc_obj <- roc(observed, cv_probs, quiet = TRUE)
auc(roc_obj) ### 0.8047; seems better than skj model

### Calibration plot shows whether the predicted probabilities are well-calibrated (i.e., when the model says 30% chance, do we actually see presence ~30% of the time?)
### Fit the calibration GAM (using out-of-fold predictions)
calib_gam <- gam(station_data$Thunnus_pa ~ s(cv_probs, bs = "ts"),
                 family = binomial(link = "logit"),
                 method = "REML")
summary(calib_gam)

### Predict smooth curve + uncertainty on a fine grid
newdata <- data.frame(cv_probs = seq(0, 1, length.out = 477))
pred <- predict(calib_gam, 
                newdata = newdata, 
                type = "link", 
                se.fit = TRUE)

### convert to probability scale
fit_prob <- plogis(pred$fit)
ci_lower <- plogis(pred$fit - 1.96 * pred$se.fit)
ci_upper <- plogis(pred$fit + 1.96 * pred$se.fit)

### Plot
#png("Thunnus_PA_calibration.png", units = "in", width = 6, height = 4, res = 1200)
plot(0, 0, 
     type = "n", 
     xlim = c(0, 1), 
     ylim = c(0, 1),
     xlab = "Predicted Probability", 
     ylab = "Observed Probability",
     las = 1, 
     bty = "l", 
     cex.main = 1.1, 
     cex.lab = 1.1)
abline(0, 1, 
       col = "gray50", 
       lty = 2, 
       lwd = 1.6)
polygon(c(newdata$cv_probs, rev(newdata$cv_probs)),
        c(ci_lower, rev(ci_upper)),
        col = rgb(0.78, 0.78, 0.78, 0.35),
        border = NA)
lines(newdata$cv_probs, 
      fit_prob, 
      col = "black", 
      lwd = 2.6)
rug(cv_probs[station_data$Skipjack_pa == 0], 
    side = 1, 
    col = "gray40", 
    lwd = 0.8, 
    ticksize = 0.025)
rug(cv_probs[station_data$Skipjack_pa == 1], 
    side = 3, 
    col = "gray20", 
    lwd = 0.8, 
    ticksize = 0.025)
#dev.off()
```

### Plotting number of each taxon in feature types
```{r}
### Starting by making a new column where each station has a classification as "front", "eddy", or "neither/oceanic"
station_data <- station_data %>%
  mutate(feature = case_when(ADCP_velo > 0.375 & SSS > 35.25 ~ "Front",
                             SST...C. > 29.5 & Chla < 7 ~ "Eddy",
                             TRUE ~ "Oceanic"))

### Boxplots
density_boxplot_data <- station_data %>%
  filter(Skipjack.Density.1000m3 > 0 | Thunnus.Density.1000m3 > 0) %>%
  pivot_longer(cols = c(Skipjack.Density.1000m3, Thunnus.Density.1000m3),
               names_to = "Genus",
               values_to = "density") %>%
  mutate(Genus = case_when(Genus == "Skipjack.Density.1000m3" ~ "Katsuwonus",
                           Genus == "Thunnus.Density.1000m3" ~ "Thunnus",
                           TRUE ~ Genus))

feature_genus_means <- density_boxplot_data %>%
  group_by(feature, Genus) %>%
  summarise(mean_density = mean(density, na.rm = TRUE),
            .groups = "drop")

#png("density_feature_nonzero.png", units = "in", width = 6, height = 5, res = 1200)
ggplot(density_boxplot_data, 
       aes(x = feature, 
           y = density, 
           fill = Genus)) +
  geom_boxplot(aes(group = interaction(feature, Genus)),
               position = position_dodge(width = 0.75),
               width = 0.70,
               color = "black",
               linewidth = 0.4,
               outlier.shape = 21) +
  geom_point(data = feature_genus_means,
             aes(x = feature, 
                 y = mean_density, 
                 group = Genus),
             position = position_dodge(width = 0.75),
             color = "black",
             size = 2,
             inherit.aes = FALSE) +
  scale_fill_manual(values = c("Katsuwonus" = "darkslategray2",
                               "Thunnus" = "goldenrod1")) +
  scale_color_manual(values = c("Katsuwonus" = "darkslategray2",
                                "Thunnus" = "goldenrod1")) +
  scale_y_continuous(limits = c(0, 15)) +
  labs(x = "Feature type",
       y = "Larval Density (ind./1,000 m³)",
       fill = "Genus",
       color = "Genus") +
  theme_minimal(base_size = 14)
#dev.off()

table(density_boxplot_data$feature)

### Tweedie GLM
glm_tweedie <- glmmTMB(density ~ Genus * feature,
                       data = density_boxplot_data,
                       family = tweedie(link = "log"))

summary(glm_tweedie)
emmeans(glm_tweedie, ~ feature | Genus, type = "response")
pairs(emmeans(glm_tweedie, ~ feature | Genus, type = "response"), adjust = "tukey")
```
